---
title: "Shelter data from FAIMS v 2.6"
author: "Adela Sobotkova"
date: "2023-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Shelter data processing from FAIMS modules

This script is preparatory and needs only to be run when new FAIMS data are collected. If you are trying to visualize existing shelters, go on to use the LoadShelters.R.

This script imports, projects and combines data
from the two FAIMS2.6 modules (online and local instantiation) from 2023 April and May data collection in Aarhus downtown. See AarhusShelterMapLisaMarie20230510.pdf for coverage
Upon careful merge (wrong order distorts timestamp), we export the data to OR for further streamlining of multivalued columns and bring back for visualisation.

Note: Must use the 0510 online dataset. if you use the 0507 dataset, it lacks 21 points from 8-10 May !


## Libraries
```{r}
library(sf)
library(mapview)
library(tidyverse)
```

## CREATE DATA FROM FAIMS SERVER
```{r local-faims}
# To load shp or geojson from FAIMS

#sh23_local <- read_sf("C:/Users/adela/Documents/Professional/Projects/AarhusShelters/data/AarhusShelters20230515Localserver/Shelter.shp")
sh23_local <- read_sf("data/Shelter-local.shp")
# Data lacks projection so project to UTM32N to faciliate analysis
st_crs(sh23_local) <- 32632

```


```{r online-faims}
# Load data from online server and process similarly 
sh23_online <- read_sf("data/Shelter-online32632.shp") 
#sh23_online <- read_sf("C:/Users/adela/Documents/Professional/Projects/AarhusShelters/data/AarhusShelters20230510online/Shelter32632.shp") # shelters0507 lacks 23 features from 8 and 10 May
```


## Check timeline and authorship
```{r}
library(lubridate)

sh23_online %>%  # 145 features
  #filter(createdBy == "Lise Søndergaard Jensen") %>% 
  mutate(day = as_date(createdAt0)) %>% 
  group_by(day, createdBy) %>% 
  tally()

sh23_local %>%   # 21 features
  #filter(createdBy == "Lise Søndergaard Jensen") %>% 
  mutate(day = as_date(createdAt0)) %>% 
  group_by(day, createdBy) %>% 
  tally()
```

## Merge the two datasets
Careful! timestamp gets messed up with reverse order)

```{r}
# check column names are identical
colnames(sh23_local)==colnames(sh23_online) 

# merge
shelters23 <- rbind(sh23_online, sh23_local)

# check data grouped by date
shelters23 %>% 
  mutate(day = as_date(createdAt0)) %>% 
  group_by(day, createdBy) %>% 
  tally()
```

## View the merged data
```{r}
# View projected data in interactive map
mapview(shelters23, cex = "Accuracy", name = "identifier" )

# What's the GPS accuracy? Beware annotations have turned some values into 'characters'
head(shelters23$Accuracy)
summary(as.numeric(shelters23$Accuracy))

# What's the accessibility?
mapview(shelters23, zcol = "Accessibl0")

# view the data
mapview(shelters23, cex = "Accuracy")

mapview(shelters23, zcol = "LanduseAr0")

```
Many columns, such as the accessibility, landuse or type need to be cleaned up in OpenRefine

## Export as csv to process in OR
```{r}
shelters23 %>% 
  st_drop_geometry() %>% 
  write_csv("data/shelters23forOR.csv")
```

